---
title: "Example Analysis"
bibliography: citations.bib
output:
  quarto::quarto_html:
---

# Women in the Workforce

## Question(s)

What is the evolution of the gender wage gap over time? How does variation in age and part-time work contribute to the gender wage gap? This exploration is inspired by some of the recent Nobel Prize award-winning work of Claudia Goldin, and attempts to replicate some of her findings using a much more limited data set [@goldin_grand_2014] [@goldin_hours_2015].

## Audience

The intended audience members for this example data analysis are other students in my statistical programming course at the Johns Hopkins Bloomberg School of Public Health. This work is a fun classroom assignment that I hope will peak the interest of other curious students.

::: callout-note
This analysis is not intended to be a serious academic piece or critique.
:::

## Data and data dictionaries

I will be using the earnings_female.csv and employed_gender.csv data sets from the 2019 TidyTuesday GitHub repo for 'Women in the Workforce' [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-03-05) [@women_workforce]. I chose not to use jobs_gender.csv since it only contains information from three years. The two data sets, earnings_female.csv and employed_gender.csv, include female percent of male salary by age group and percent of women/men working full-time/part-time from 1979 to 2011. The data originate from the Bureau of Labor and Census Bureau.

See data dictionaries [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-03-05) [@women_workforce].

## Analysis

Download data:
```{r}
#| label: "setup"
#| output: false
# Load libraries
library(dplyr)
library(ggplot2)
library(here)

# Read in data
# tuesdata <- tidytuesdayR::tt_load("2019-03-05")
# earnings_female <- tuesdata$earnings_female
# employed_gender <- tuesdata$employed_gender

```
::: callout-tip
Avoid API limit for knitting the document using the code below.
:::
```{r}
# Tests if a directory named "data" exists locally
# if (!dir.exists(here("data"))) {
#     dir.create(here("data"))}
# 
# ## Saves data only once (not each time you knit a R Markdown)
# if (!file.exists(here("data", "earnings_female.RDS"))) {
#     ### Save the file to RDS objects
#     saveRDS(earnings_female, file = here("data", "earnings_female.RDS"))
#   saveRDS(employed_gender, file = here("data", "employed_gender.RDS"))}
# 
# if (!file.exists(here("data", "employed_gender.RDS"))) {
#     ### Save the file to RDS objects
#    saveRDS(employed_gender, file = here("data", "employed_gender.RDS"))}


# Read in the .RDS dataset locally from our computing environment
earnings_female <- readRDS(here("data", "earnings_female.RDS"))
employed_gender <- readRDS(here("data", "employed_gender.RDS"))
```

Clean/wrangle data:
```{r}
# Rename 'percent' column to 'percent_female' and 'group' to 'age_group' and 'Year' to 'year' for clarity and consistency in earnings_female.csv
earnings_female <- earnings_female %>%
  rename(percent_female = percent, age_group = group, year=Year)

# Relabel 'Total, 16 years and older' to 'Total' in 'age_group' column. Remove anytime 'years' appears in age_group column and relabel '65 years and older' to '65+'
earnings_female <- earnings_female %>%
  mutate(age_group = gsub("Total, 16 years and older", "Total", age_group)) %>%
  mutate(age_group = gsub(" years", "", age_group)) %>%
  mutate(age_group = gsub("65 and older", "65+", age_group))

# Subset employed_gender.csv from 1979 to 2011 to match the time frame from the earnings_female.csv dataset
employed_gender <- employed_gender %>%
  filter(year >= 1979, year <= 2011)

# Merge two datasets by year
combined_data <- earnings_female %>%
  inner_join(employed_gender, by = "year")

# Keep only part-time % columns since the full-time % columns are redundant: full-time % = 100% - part-time. The data do not contain percentage unemployed
clean_data <- combined_data %>%
  select(year, age_group, percent_female, total_part_time, part_time_female, part_time_male) 

# View clean_data
head(clean_data)
```

First, let's explore the evolution of female percentage of male salary over time, overall and faceted by age group. Claudia Goldin's work suggests that the gender wage gap has experienced a large convergence toward equality in the last century, but difficulties for women returning to work after having children may result in reduced work in middle age [@goldin_grand_2014]. What do the data show?
```{r}
# Create line plot of female percentage of male salary over time, faceted by age group
ggplot(clean_data, aes(x = year, y = percent_female,color = age_group, group = age_group)) +
  geom_line() +
  labs(
    title = "Female salary % of male salary from 1979-2011",
    subtitle = "Women 34+ years of age have lower female salary % of male salary than other age groups",
    x = "Year",
    y = "Female salary % of male salary",
    caption = "Source: TidyTuesday 2019 'Women in the Workforce' GitHub repository"
  ) +
  scale_color_discrete(name = "Age Group") +
  theme_minimal() +
  facet_wrap(~age_group, scales = "free_y")
```
Overall, female salary percentage of male salary has risen over the years, from <65% to >80% between 1979 and 2011. But we can see that even by 2011, female salary percentage of male salary drops from >90% to <80% after 35 years of age. The data supports the idea that women experience a greater gender wage gap in the middle age and beyond.

Research from the Pew Research Center, shown on the left, supports the ideas conveyed above as well [@kochhar_enduring_2023].

![Source: [@kochhar_enduring_2023]](PewResearch.jpg){width="400" height="400"} 


Now let's take a look at percentage of part-time working females and males over time. 
```{r}
# Subset to all ages combined
clean_data_total <- clean_data %>%
  filter(age_group == "Total")
```

::: column-margin
We cannot stratify results by age groups, since the TidyTuesday data only contain percentage of part-time work females by year.
:::

```{r}
# Create scatter plot with regression line of percentage of part-time working females over time
ggplot(clean_data_total, aes(x = year, y = part_time_female)) +
  geom_point() +
  geom_smooth(method = "lm", formula = 'y~x') + 
  labs(
    title = "Percentage of part-time working females from 1979-2011",
    subtitle = "Percentage of part-time working females only decreased slightly over time",
    x = "Year",
    y = "% part-time working females",
    caption = "Source: TidyTuesday 2019 'Women in the Workforce' GitHub repository"
  ) +
  theme_minimal() 

# Create line plot of percentage of part-time working males over time
ggplot(clean_data_total, aes(x = year, y = part_time_male)) +
  geom_point() +
  geom_smooth(method = "lm", formula = 'y~x') + 
  labs(
    title = "Percentage part-time working males from 1979-2011",
    subtitle = "Percentage of part-time working males increased slightly over time",
    x = "Year",
    y = "% part-time working males",
    caption = "Source: TidyTuesday 2019 'Women in the Workforce' GitHub repository"
  ) +
  theme_minimal()
```
The regression line shows that percentage of part-time working females fluctuated from 1979 to 2011, but decreased only slightly (by 1 percentage point).  Percentage of part-time working males from 1979 to 2011 only increased by 2 percentage points. These mild results hint that the reduction of the gender wage gap that occurred during this time span may not be explained by changes in part-time work patterns, but rather other societal and lifestyle factors. In fact, Claudia Goldin's work points to part-time work as one of the persisting contributors to the remaining wage gap [@goldin_hours_2015].

## Summary

In summary, the gender wage gap greatly reduced between 1979 to 2011 from <65% to >80%. Female salary percentage of male salary drops after 35 years of age, and this decreased is still pronounced by 2011. Percentage of part-time working females only decreased by 1 percentage point and percentage part-time working males only increased by 2 percentage points between 1979 to 2011. Perhaps more nuanced historical data on other factors such as employer bias would benefit this currently limited analysis by providing a clearer picture for the reasons behind the reduction of the gender wage gap. 

## Functions from dplyr, ggplot2

dplyr functions: 'rename', 'mutate', 'filter', 'inner_join', and 'select' 

ggplot2 geom_* functions: geom_line, geom_point, geom_smooth

## References
